Bayesian Rate Estimation

### 2012-11-10 many.tex
Section re-org.

This paper is organized as follows.  In section \ref{sec:model} we
introduce our method for Bayesian rate estimation, and describe
loudest-event and foreground-dominated statistics for comparison.
Section \ref{sec:GW-example} includes three examples of the
application of these techniques.  We conclude with a discussion of
important remaining issues in section \ref{sec:discussion}.

@@

@@

\subsubsection{Loudest event statistic}
If we were to include only the $k$ loudest events in the posterior
distribution, rather than all observed events, the posterior
(Eq.~\eqref{eq:posterior}) would be modified by an additional factor
of $\exp[R_f \hat{F}(x_{N-k+1},\theta) + R_b
  \hat{B}(x_{N-k+1},\theta)]$, where we have assumed events are
ordered by loudness, so that $x_{N-k+1}$ is the $k$-th loudest
event. This term accounts for the data-dependent threshold that a
loudest event statistic employs.

@@

\subsubsection{Foreground dominated statistic}
If we set the threshold for including an event, $x_{\rm min}$,
sufficiently high, we can ensure that $\hat{f}(x_i,\theta) \gg
\hat{b}(x_i,\theta)$ for all ranking statistics $x_i$ in the data
set. The posterior can then be approximated by
\begin{equation}
p_{\rm FD}\left( R_f, R_b, \theta | d \right) \propto \prod_{i} \left[
  \hat{f}\left(x_i, \theta\right)\right] R_f^N \exp\left[-\left( R_f +
  R_b \right) \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{equation}
Normalisation over $R_b$ gives a constant factor and the posterior on
the foreground rate becomes
\begin{equation}
p_{\rm FD}\left( R_f, \theta | d \right) \propto \prod_{i} \left[
  \hat{f}\left(x_i, \theta\right)\right] R_f^{N-\frac{1}{2}}
\exp\left[-R_f \right] p(\theta).
\end{equation}
Ignoring the dependence on $\theta$, this is peaked at a rate $R_f =
N-1/2$, so we have the expected result that, in the foreground
dominated regime, the rate is approximately equal to the number of
events observed (the $1/2$ comes from our use of the Jeffreys prior on
the rate).  We note that the rate $R_f$ is defined as the rate of
events occurring above the specified threshold and therefore if the
foreground dominated statistic were used to estimate rates $R_{f,1}$
and $R_{f,2}$ for thresholds $x_{\rm min}=x_1$, $x_{\rm min}=x_2$
respectively, these rates should be compared by equating $R_{f,1}
(1-\hat{F}(x_2,\theta))$ and $R_{f,2} (1-\hat{F}(x_1,\theta))$.

@@

\subsection{Dependence of the Bayesian Rate Estimate on the Value of the Threshold}
This paper is concerned with Bayesian rate estimates based on a lists
of events (which are a certain kind of summary of the complete data
set): either the loudest event, or a list of all events whose SNR
exceeds a certain, pre-set threshold.  In this subsection we address
the question of how the estimate depends on the threshold value.

For Fig.~\ref{fig:delRb}, we adopted a flat prior on $R$ between $1$
and $10$ (i.e, we took $p(R) = 1/9$ for $1 < R <10$, and zero outside
that interval).
% To create Fig.~\ref{fig:delR1}, 
We drew $10^4$ random realizations $(R_t, data)$, where $R_t$ was
drawn from that flat prior and the data--a list of $\rho$ values-- was
drawn from the distribution $p(data|R_t)$ given by Eq.~\ref{eq:ptot} .
For any particular threshold value, $\rho_{th}$, we created an event
list by taking the subset of the $\rho$ values
% in the data set that are 
that were greater than $\rho_{th}$.  For each data set, we considered
a sequence of increasing threshold values $\rho_{th}$, using them to
create a corresponding sequence of (ever shorter) event lists.  For
each event list, we calculated $\Delta R$ and $R_B - R_t$.
%and take the mean over all realizations.  
As a code check, we verified that the mean of $R_B - R_t$ (over all
realizations) was consistent with zero for all $\rho_{th}$.
Fig.~\ref{fig:delR1} displays both the mean value of $\Delta R$ and
the mean number of noise events (as opposed to actual signals) as a
function of $\rho_{th}$.  Note the relative insensitivity of $\Delta
R_t$ to $\rho_{th}$: as the mean number of noise events on the event
list shrinks from $8$ to $\sim 10^{-6}$, the average $\Delta R_B$
increases only $\sim 14\%$, from $0.42$ to $0.48$.

>>>>

@@

\section{Comparison to Other Rate Estimation Methods}

@@

\subsection{Loudest event statistic}
If we were to include only the $k$ loudest events in the posterior
distribution, rather than all observed events, the posterior
(Eq.~\eqref{eq:posterior}) would be modified by an additional factor
of $\exp[R_f \hat{F}(x_{N-k+1},\theta) + R_b
  \hat{B}(x_{N-k+1},\theta)]$, where we have assumed events are
ordered by loudness, so that $x_{N-k+1}$ is the $k$-th loudest
event. This term accounts for the data-dependent threshold that a
loudest event statistic employs.

@@

\subsection{Foreground dominated statistic}
If we set the threshold for including an event, $x_{\rm min}$,
sufficiently high, we can ensure that $\hat{f}(x_i,\theta) \gg
\hat{b}(x_i,\theta)$ for all ranking statistics $x_i$ in the data
set. The posterior can then be approximated by
\begin{equation}
p_{\rm FD}\left( R_f, R_b, \theta | d \right) \propto \prod_{i} \left[
  \hat{f}\left(x_i, \theta\right)\right] R_f^N \exp\left[-\left( R_f +
  R_b \right) \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{equation}
Normalisation over $R_b$ gives a constant factor and the posterior on
the foreground rate becomes
\begin{equation}
p_{\rm FD}\left( R_f, \theta | d \right) \propto \prod_{i} \left[
  \hat{f}\left(x_i, \theta\right)\right] R_f^{N-\frac{1}{2}}
\exp\left[-R_f \right] p(\theta).
\end{equation}
Ignoring the dependence on $\theta$, this is peaked at a rate $R_f =
N-1/2$, so we have the expected result that, in the foreground
dominated regime, the rate is approximately equal to the number of
events observed (the $1/2$ comes from our use of the Jeffreys prior on
the rate).  We note that the rate $R_f$ is defined as the rate of
events occurring above the specified threshold and therefore if the
foreground dominated statistic were used to estimate rates $R_{f,1}$
and $R_{f,2}$ for thresholds $x_{\rm min}=x_1$, $x_{\rm min}=x_2$
respectively, these rates should be compared by equating $R_{f,1}
(1-\hat{F}(x_2,\theta))$ and $R_{f,2} (1-\hat{F}(x_1,\theta))$.

\section{Dependence of the Bayesian Rate Estimate on the Value of the Threshold}
This paper is concerned with Bayesian rate estimates based on a lists
of events (which are a certain kind of summary of the complete data
set): either the loudest event, or a list of all events whose SNR
exceeds a certain, pre-set threshold.  In this subsection we address
the question of how the estimate depends on the threshold value.

For Fig.~\ref{fig:delRb}, we adopted a flat prior on $R$ between $1$
and $10$ (i.e, we took $p(R) = 1/9$ for $1 < R <10$, and zero outside
that interval).
% To create Fig.~\ref{fig:delR1}, 
We drew $10^4$ random realizations $(R_t, data)$, where $R_t$ was
drawn from that flat prior and the data--a list of $\rho$ values-- was
drawn from the distribution $p(data|R_t)$ given by Eq.~\ref{eq:ptot} .
For any particular threshold value, $\rho_{th}$, we created an event
list by taking the subset of the $\rho$ values
% in the data set that are 
that were greater than $\rho_{th}$.  For each data set, we considered
a sequence of increasing threshold values $\rho_{th}$, using them to
create a corresponding sequence of (ever shorter) event lists.  For
each event list, we calculated $\Delta R$ and $R_B - R_t$.
%and take the mean over all realizations.  
As a code check, we verified that the mean of $R_B - R_t$ (over all
realizations) was consistent with zero for all $\rho_{th}$.
Fig.~\ref{fig:delRb} displays both the mean value of $\Delta R$ and
the mean number of noise events (as opposed to actual signals) as a
function of $\rho_{th}$.  Note the relative insensitivity of $\Delta
R_t$ to $\rho_{th}$: as the mean number of noise events on the event
list shrinks from $8$ to $\sim 10^{-6}$, the average $\Delta R_B$
increases only $\sim 14\%$, from $0.42$ to $0.48$.

@@
